---
title: "Titanic: dead or alive"
output:
  html_document:
    toc: false
    theme: flatly
---

```{r include=FALSE, message=FALSE, warning=FALSE}
library(car)
library(ggplot2)
library(gridExtra)
library(dplyr)
```

In this tutorial we use **logistic regression** to predict (model) whether passengers on the Titanic lived or died, as determined by passenger class (1st, 2nd, 3rd), age and sex (male, female).

First we read the data, and convert 'Survived' to a variable with more meaningful labels:

```{r}
library(lgrdata)
data(titanic)

titanic$Survived <- factor(ifelse(titanic$Survived == 1, "yes", "no"))
```

Let's look at a few random rows,

```{r}
library(dplyr)
sample_n(titanic, 8)
```


The idea here is to find out whether the probability of survival depends on the passenger's `Age`, `Sex` and `PClass` (passenger class). Before we proceed, it is always a good idea to start by visualizing the data to find out what we are dealing with (and to make sure we will interpret the model output correctly). If we plot two factor variables against each other, R produces a 'mosaic plot', as the following example demonstrates.

```{r facfacplot, fig.cap="Probability of survival versus passenger class and sex for the titanic data."}
par(mfrow=c(1,2), mgp=c(2,1,0))
with(titanic, plot(Sex, Survived, ylab="Survived", xlab="Sex"))
with(titanic, plot(PClass, Survived, ylab="Survived", xlab="Passenger class"))
```


## Logistic regression

In a 'normal' linear regression model, we use `lm` to study the dependence of a *continuous response* variable on one or more *predictors*. If the response variable is not continuous, but instead takes only two values (failure and success, 0 or 1, and so on), we must use logistic regression to study the dependence.

In R you can use `glm` to achieve this, which is used in pretty much the same way as `lm`. Also, you have to add the argument `family=binomial` to indicate logistic regression, as `glm` is a more general function that allows other models.


```{r}
fit_1 <- glm(Survived ~ Age + Sex + PClass, data=titanic, family=binomial)
```


The first step is to check whether all predictors are *significant*, that is, whether there is evidence that they contribute to explaining variation in the survival status of the passengers.

Here we prefer the `Anova` function from the `car` package. Small p-values (indicated with `***`) mean the factors are significant.

```{r}
library(car)
Anova(fit_1)
```

It seems all three variables are somehow "significant" in the model, but so what? We are usually more interested in **how** the relationships vary. 

We have fit a model with all three variables, and to make sense of it we use the `visreg` package (**vis**ualisation of **reg**ression models).

We start with looking at the effect of one variable, `Age`.

```{r}
library(visreg)

visreg(fit_1, "Age", scale = "response")
```

Here, the plot shows the effect of passenger age (X-axis) on the probability that a passenger survived. We use `scale = "response"` especially for `glm` models, to make sure we plot a probability on the Y-axis.

The above plot is constructed by keeping all other variables out of the model, and only visualizing what happens when you vary one of them. We can add another variable with the `by` argument, like so:

```{r}
visreg(fit_1, "Age", by = "PClass", overlay = TRUE, scale = "response")
```
 
Here we also like to use `overlay = TRUE`, otherwise we end up with three panels (in this plot, the effects are easier to study).

We now conclude that younger passengers were (much!) more likely to survive, and first class passengers more likely than lower classes. 

What about male vs. female passengers? First, we look at the difference between male and female passengers by passenger class, holding the age of the passenger constant. To do this, we simply leave out age in `visreg` - the default behaviour is to set age to the mean across all data points.

Also, here we use `gg=TRUE` to make a `ggplot2` figure, which allows use of a theme (and whatever `ggplot2` settings you like):

```{r}
library(ggplot2)
visreg(fit_1, "Sex", by="PClass", gg=TRUE) + 
  theme_minimal()
```

There seems to be a large difference between male and female survivorship. We can make a different plot with a small adjustment:

```{r}
library(ggplot2)
visreg(fit_1, "PClass", by="Sex", gg=TRUE, scale = "response") + 
  theme_minimal()
```

Now the difference between male and female passengers is very striking - as is the difference between passenger classes.

Finally, we can make a pair of plots, visualizing all three variables and their effect on passenger survival. As earlier, we make a plot of survival probability by passenger age and passenger class, one for each of 'female' and 'male' passengers.

With the use of the `cond` argument, we can take subsets of the model predictions. Also note the use of `ggExtra` to place the two plots side by side. We also use settings `rug=FALSE` to switch off the vertical lines indicating observations, and add a couple of small `ggplot2` settings (title, theme).

```{r}
library(gridExtra)

plot_male <- visreg(fit_1, "Age", 
                    by="PClass", 
                    scale = "response",
                    overlay = TRUE,
                    gg = TRUE,
                    rug = FALSE,
                    cond=list(Sex="male")) + 
  theme_minimal() +
  ylim(0,1) +
  theme(legend.position = "none") +
  labs(title = "Men")

plot_female <- visreg(fit_1, "Age", 
                    by="PClass", 
                    scale = "response",
                    overlay = TRUE,
                    gg = TRUE,
                    rug = FALSE,
                    cond=list(Sex="female")) +
  theme_minimal() +
  ylim(0,1) +
  theme(legend.position = "none") +
  labs(title = "Women")

grid.arrange(plot_male, plot_female, ncol = 2)
```














